---
title: 'R Notebook sandbox: IMDB data load (Sept 2020)'
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
    number_sections: true
my-var: "monte"  # https://bookdown.org/yihui/rmarkdown/html-document.html
---



# Setup


### source setup
```{r}
github.monte.raw = "https://raw.githubusercontent.com/MonteShaffer/";
include.setup = paste0(github.monte.raw, "humanVerse/main/include.setup.R");
source(include.setup);  # Maybe comment this out, if error 
## TODO, figure out what functions need to go in there ...
 
github.monte.http = "https://github.com/MonteShaffer/";
######## we will parse this page to get a list of the .R functions to include ########
github.monte.humanVerse = paste0(github.monte.http, "humanVerse/tree/main/humanVerse/R/"); 

######## you can pass flag `force.download = TRUE` if you want to make certain it is not coming from cache ########

# getSourceLocation();
# Sys.setenv("HUMANVERSE_CACHE" = "R:/R-humanVerse-cache"); 
# getSourceLocation();
###### R::humanVerse #####
includeGithubFolder(github.monte.humanVerse, force.download = TRUE); 

# github.includeFolder(github.monte.humanVerse, force.download = TRUE); 


```

```{r}
github.monte.humanVerse.misc = paste0(github.monte.http, "humanVerse/tree/main/misc/"); 

includeGithubFolder(github.monte.humanVerse.misc, force.download = TRUE); 

# github.includeFolder(github.monte.humanVerse.misc, force.download = TRUE); 

```

### load imdb data from dataWar

```{r}
main = "https://github.com/DataWar/imdb/";
raw = "https://raw.githubusercontent.com/DataWar/imdb/";

imdb.data = data.load("2020-Sept", "imdb", mode="rds", main=main, raw=raw, sub="");

# IMDB.loadData();
```

It is globalized and the "older" functions will work.
```{r}
.GlobalEnv$imdb.data = imdb.data;
```

### load rank data

This used eigen rank.  There is no MONEY tied to these rankings, merely, who worked with whom.  Top-15 cast.  I do **NOT** think it included the top-creatives: writers/directors.  

Top-50 movies per year were included.
<https://www.imdb.com/search/title/?year=1973&title_type=feature&>

Modern "Popularity" may be a bad "seed" to do the analysis.

```{r}
my.rank = list();
my.rank$movies = list();

my.rank$movies$alltime = data.load("2020-Sept", "movieRank5000", mode="rds", main=main, raw=raw, sub="");
my.rank$movies$since1980 = data.load("2020-Sept", "movieRank2000", mode="rds", main=main, raw=raw, sub="");


my.rank$actors = list();

my.rank$actors$alltime = data.load("2020-Sept", "actorRank5000", mode="rds", main=main, raw=raw, sub="");
my.rank$actors$since1980 = data.load("2020-Sept", "actorRank2000", mode="rds", main=main, raw=raw, sub="");

```


```{r}
View(my.rank$actors$since1980)
```

```{r}
View(my.rank$actors$alltime)
```


```{r}
View(my.rank$movies$since1980)
```

```{r}
View(my.rank$movies$alltime)
```

# Will is Better

```{r}

IMDB.searchPersonName("Will* Smith", 1:3);
```

```{r}
will = list();
will$nmid = "nm0000226";
will$movies = IMDB.getMoviesForPerson(will$nmid);
will$ttids = will$movies$ttid;

```



# Social Network

```{r}
network = imdb.data$movies.df$cast; 
names(network);
dim(network);

```

## Top 250

```{r}
res = IMDB.buildSocialNetwork(ttids = imdb.data$movies$top250, use.cpp=FALSE);

names(res);


```

```{r}
top250MovieRank = matrix.computeEigenRank(res$MM);

dim(res$MM);

top250MovieRank$ttid = top250MovieRank$nodes;

top250MovieRank = merge(top250MovieRank, imdb.data$all.actors.movies, by=c("ttid") );

top250MovieRank;
#View(top250MovieRank);
```


```{r}
top250ActorRank = matrix.computeEigenRank(res$AA);

dim(res$AA);

top250ActorRank$nmid = top250ActorRank$nodes;

top250ActorRank = merge(top250ActorRank, imdb.data$all.actors.info, by=c("nmid") );

top250ActorRank;
#View(top250ActorRank);
```


## Will Smith
```{r}

res = IMDB.buildSocialNetwork(ttids = will$movies$ttid, use.cpp=FALSE);

names(res);
```

```{r}

willMovieRank = matrix.computeEigenRank(res$MM);
willMovieRank;
```


```{r}

willActorRank = matrix.computeEigenRank(res$AA);
willActorRank;

```

