---
title: 'R Notebook sandbox: Protein in Diet'
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
    number_sections: true
my-var: "monte"  # https://bookdown.org/yihui/rmarkdown/html-document.html
---



# Setup


### source setup
```{r}
github.monte.raw = "https://raw.githubusercontent.com/MonteShaffer/";
include.setup = paste0(github.monte.raw, "humanVerse/main/include.setup.R");
source(include.setup);  # Maybe comment this out, if error 
 
github.monte.http = "https://github.com/MonteShaffer/";
######## we will parse this page to get a list of the .R functions to include ########
github.monte.humanVerse = paste0(github.monte.http, "humanVerse/tree/main/humanVerse/R/"); 

######## you can pass flag `force.download = TRUE` if you want to make certain it is not coming from cache ########

# getSourceLocation();
###### R::humanVerse #####
includeGithubFolder(github.monte.humanVerse, force.download = TRUE); 
```

# Data Provenance

How good is your data?  Have you documented your data?

## Protein

Here is another dataset often used in multivariate statistics courses.

- <https://github.com/MonteShaffer/humanVerse/blob/main/data/protein/protein.md>

Notice that the information is a bit more detailed, and it has been published in at least three locations.  Although we don't know exactly how the data was collected and "averaged", we can "assume" that some reasonable protocols were estaplished.  The earliest publication was Weber (1973), so we would assume that the data was collected circa 1970.

I would give this dataset a reasonably good score on the data-provenance scale, maybe a 9/10.  Everything looks good, and that I don't have more detail may likely be a function of the era (1970s).

I think about this data-provenance quality like a time capsule.  If someone discovered your data and notebooks in a distant future time, could they understand and recreate what you have prepared and analyzed.


### Data Import

#### Text Import
```{r}
data.file = paste0(github.monte, "humanVerse/main/data/protein/protein.txt");  

protein.txt = readFromPipe(data.file);
# notice, it did not like a column name starting with a number.
```

#### RDS Import
```{r}
data.file = paste0(github.monte, "humanVerse/main/data/protein/protein.rds");  

protein.rds = readRDS(url(data.file));
# notice, this binary form didn't care about the column name format ...
```

We don't know the exact units of the proteins, but with a bit more research, we could likely find that scale (if you can read German).  I wonder if the 1973 article was published in East or West Germany?  Is the USA in the dataset?

### hclust

#### Country-level Analysis
```{r}
protein.df = protein.rds;
X = as.matrix(protein.df[,c(2:10)]) ;
  rownames(X) = protein.df$Country;
  
Xs = scale(X);
# you can change which X is being used: X or Xs
whichX = X;
```

```{r}
X.d = dist(whichX);  # Euclidean Distance

X.hclust = hclust(X.d);  # method = "complete"
plot(X.hclust, main="Countries (complete)");
```

```{r}
X.hclust = hclust(X.d, method="ward.D2");
plot(X.hclust, main="Countries (ward.D2)");
```

##### Cutting the Tree into Clusters
```{r}
numberClusters = 8;

X.cut = stats::cutree(X.hclust, k=numberClusters);
table(X.cut);
X.cut;
```

##### Subtrees
```{r}

library(dendextend);
colors = colorspace::rainbow_hcl(numberClusters);

dend = stats::as.dendrogram(X.hclust);
dend = dendextend::color_branches(dend, k = numberClusters);

  graphics::par(mfrow = c(1,1));
  graphics::plot(dend, main="Countries (ward.D2)", ylim=c(0, ceiling(max(X.hclust$height))) );  # Can I color-code the labels?  "Fe"

# maybe make custom ceiling approach that rounds to a significant digit ... 76425 -> 77000 or 76500
```

```{r}

dend.labels = base::labels(dend);
  groups = dendextend::cutree(dend, k=numberClusters, order_clusters_as_data = FALSE);

graphics::par(mfrow = c(2,2));

dends = list();
for(i in 1:numberClusters)
  {
  keep.me    = dend.labels[i != groups];
  dends[[i]] = dendextend::prune(dend, keep.me);
  graphics::plot(dends[[i]], cex=0.5, main = paste0("SubTree number ", i) );  # , ylim=c(0,100)
  }

# restore plot
  graphics::par(mfrow = c(1,1));
  
```

#### Protein-level Analysis
```{r}
X = as.matrix(protein.df[,c(2:10)]) ;
  rownames(X) = protein.df$Country;
X = transposeMatrix(X);

Xs = scale(X);
# you can change which X is being used: X or Xs
whichX = X;
```

```{r}
X.d = dist(whichX);  # Euclidean Distance

X.hclust = hclust(X.d);  # method = "complete"
plot(X.hclust, main="Protein (complete)");
```

```{r}
X.hclust = hclust(X.d, method="ward.D2");
plot(X.hclust, main="Protein (ward.D2)");
```

##### Cutting the Tree into Clusters
```{r}
numberClusters = 4;

X.cut = stats::cutree(X.hclust, k=numberClusters);
table(X.cut);
X.cut;
```

##### Subtrees
```{r}

library(dendextend);
colors = colorspace::rainbow_hcl(numberClusters);

dend = stats::as.dendrogram(X.hclust);
dend = dendextend::color_branches(dend, k = numberClusters);

  graphics::par(mfrow = c(1,1));
  graphics::plot(dend, main="Protein (ward.D2)", ylim=c(0, ceiling(max(X.hclust$height))) );  # Can I color-code the labels?  "Fe"

```

```{r}

dend.labels = base::labels(dend);
  groups = dendextend::cutree(dend, k=numberClusters, order_clusters_as_data = FALSE);

graphics::par(mfrow = c(2,2));

dends = list();
for(i in 1:numberClusters)
  {
  keep.me    = dend.labels[i != groups];
  dends[[i]] = dendextend::prune(dend, keep.me);
  graphics::plot(dends[[i]], cex=0.5, main = paste0("SubTree number ", i) );  # , ylim=c(0,100)
  }

# restore plot
  graphics::par(mfrow = c(1,1));
  
```

### PCA

```{r}
X = as.matrix(protein.df[,c(2:10)]) ;
  rownames(X) = protein.df$Country;
# X = transposeMatrix(X);

Xs = scale(X);

Xs.PCA = stats::prcomp(Xs);
Xs.SVD = base::svd(Xs);
VAF = summary(Xs.PCA)$importance[2,];
```

#### Biplots
```{r}
biplot(Xs.PCA, 1:2, xlab=paste0("PC1: (", 100 * VAF[1], "%)"), ylab=paste0("PC2: (", 100 * VAF[2], "%)"));
```


```{r}
biplot(Xs.PCA, 3:2, xlab=paste0("PC3: (", 100 * VAF[3], "%)"), ylab=paste0("PC2: (", 100 * VAF[2], "%)"));
```

```{r}
biplot(Xs.PCA, 3:4, xlab=paste0("PC3: (", 100 * VAF[3], "%)"), ylab=paste0("PC4: (", 100 * VAF[4], "%)"));
```
```{r}
biplot(Xs.PCA, 5:4, xlab=paste0("PC5: (", 100 * VAF[5], "%)"), ylab=paste0("PC4: (", 100 * VAF[4], "%)"));
```
```{r}
biplot(Xs.PCA, 5:6, xlab=paste0("PC5: (", 100 * VAF[5], "%)"), ylab=paste0("PC6: (", 100 * VAF[6], "%)"));
```

```{r}
biplot(Xs.PCA, 7:6, xlab=paste0("PC7: (", 100 * VAF[7], "%)"), ylab=paste0("PC6: (", 100 * VAF[6], "%)"));
```

```{r}
biplot(Xs.PCA, 7:8, xlab=paste0("PC7: (", 100 * VAF[7], "%)"), ylab=paste0("PC8: (", 100 * VAF[8], "%)"));
```

```{r}
biplot(Xs.PCA, 9:8, xlab=paste0("PC9: (", 100 * VAF[9], "%)"), ylab=paste0("PC8: (", 100 * VAF[8], "%)"));
```


# Conclusion 

More meaningful interpretation comes from better data inputs.
