---
title: 'R Notebook sandbox: Playing with Correlation (owell)'
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
    number_sections: true
my-var: "monte"  # https://bookdown.org/yihui/rmarkdown/html-document.html
---


# Setup

I will be using this data with all of its variables often, so I can place all those chunks in a "local" setup-well.R file and just include it.

### source setup
```{r}
source("setup-owell.R");
```

We have lots of variables.  All of the above gets built automatically.  If we need to tweak something, we could tweak the one file above, and we may have 5 "personal workspace" notebooks that are accessing it.

# Correlation

Correlations are formed in a pair-wise manner, comparing a variable X to Y.

```{r}
names(wells.df);

X = removeColumnsFromDataFrame(wells.df, c("geology", "Wilcox")); 
                  # a matrix needs numbers.
                  # remove other columns that are repeative or not interesting ?
Xs = scale(X);

options(digits = 3);
X.cor = as.data.frame( cor(X) );
X.cor;
View(X.cor); # open in detached window
```

Notice we can now see relationships between the "geology" because the data is now numeric.

## pH

"At 25 °C, solutions with a pH less than 7 are acidic, and solutions with a pH greater than 7 are basic. Solutions with a pH of 7 at this temperature are neutral (e.g. pure water)."  <https://en.wikipedia.org/wiki/PH>

Notice pH is a decimal "log10" scale.

```{r}
10^7; log10(10^7);
```
6.9 means more "basic" (having base) and 7.1 means more "acidic" (having acid).

### pairwise correlations

Default is "pearson" ... 
```{r}
# pH is 28 in my dataframe
# names(X.cor);
X.cor = as.data.frame( cor(X) );
X.cor[,c(1,28)];

```

<https://sciencing.com/salt-affect-ph-water-5155101.html>

Notice "Na" and its relationship it "pH" ... is that expected?

What other interesting things do you notice?


### Significance
A hypothesis test can assess whether or not a correlation between two variables is "significant" with the given sample of data.  Recall, like most such tests, the size of the sample will influence the significance.  We have a smallish-small (n=23).
```{r}
# pH is 28 in my dataframe
# names(X.cor);
names(X.cor)[c(28,34)];
cor.test( X[, 28], X[, 34] );  # a single test
```

#### Library Hmisc

```{r}
library(Hmisc); # install.packages("Hmisc");
    rcorr(as.matrix(X), type="pearson");

```

TMI or TL;DR

#### Visualizing
```{r}
library(corrgram); # install.packages("corrgram",dependencies=TRUE);
    corrgram(X, 
              order=TRUE, 
              lower.panel=panel.shade,
              upper.panel=panel.pie,
              text.panel=panel.txt,
              main="My title");


```

#### Causation vs Dependence



Does zero correlation imply independence?

```{r, chunck-correlation-causation}
## Correlation does not imply causation 


X = rnorm(100,0,1);
  Y = ( X - mean(X) ) / sd(X);
cor(X,Y);  ## Linear combinations

  Y = X^2;
cor(X,Y);  ## Non-linear relationship

  Y = ( sample(X) - mean(X) ) / sd(X);
cor(X,Y);  ## Randomized relationship [sample mixes up X ?]
```

