---
title: 'R Notebook: Playing with Distance (STORE LOCATOR)'
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: true
    number_sections: true
my-var: "monte"  # https://bookdown.org/yihui/rmarkdown/html-document.html
---

# Setup

## include-setup.R

This one file has all the basic functions we need to load more functions.

### source setup
```{r}
github.monte.raw = "https://raw.githubusercontent.com/MonteShaffer/";
include.setup = paste0(github.monte.raw, "humanVerse/main/include.setup.R");
source(include.setup);
```

### Custom Sourcing

#### TEMP folder
```{r}
getSourceLocation();
```

#### source (humanVerse/R)
```{r}
github.monte.http = "https://github.com/MonteShaffer/";
######## we will parse this page to get a list of the .R functions to include ########
github.monte.humanVerse = paste0(github.monte.http, "humanVerse/tree/main/humanVerse/R/"); 

######## you can pass flag `force.download = TRUE` if you want to make certain it is not coming from cache ########

###### R::humanVerse #####
includeGithubFolder(github.monte.humanVerse, force.download = TRUE); 
```

#### source (humanVerse/misc)
```{r}
######## we will parse this page to get a list of the .R functions to include ########
github.monte.misc = paste0(github.monte.http, "humanVerse/tree/main/misc/"); 

######## you can pass flag `force.download = TRUE` if you want to make certain it is not coming from cache ########

###### R::humanVerse #####
includeGithubFolder(github.monte.misc, force.download = TRUE);
```


# Store Locator
I did some work for a company called 'organicgirl' a few years back.  We built a store locator as part of their brand presence online.  It was in production for about 8 years.  You can check out their "updated version" that doesn't full operate.

The idea is that you input a ZIP code, we have a database that loosely maps that ZIP code to a latitude and longitude.  Then you query a database to find other entities (e.g., stores) within a given radius.

Searching for a radius in a database is expensive.  So it is easier to search on a square.  After SQL gives you the result for a square, you manually compute the distances of each to the query input and reduce the result said to the circle (the inscribed circle in the square).

So I created a SQL sandbox and have populated it with a table of ZIPCODES for the USA and CANADA.  It has 864,000 records (most are from CANADA oddly enough).  I removed CANADA, and I believe the US-data is intack.  About 42,000 records.

To keep the database connection information "PRIVATE", please see the `_SECRET_database_.txt` file in the DROPBOX (in the `_SECRET_` folder).  This is an example of some data-provenance constraints you may have.  It should not go directly into your GITHUB workspace as it should not be uploaded to the internet.  It's a "SECRET"...

## Load SECRET DATABASE CONNECTION information
```{r}
.db.host   = Sys.getenv("WSU_SANDBOX_HOST");
.db.name   = Sys.getenv("WSU_SANDBOX_DATABASE");
.db.user   = Sys.getenv("WSU_SANDBOX_USER");
.db.passwd = Sys.getenv("WSU_SANDBOX_PASSWD");

# a variable with a .variable (dot at the front) will be slightly invisible
```

## Prepare SQL query

```{r}
library(RMySQL); # install.packages("RMySQL");

mysql.query.template = "SELECT * FROM {tablename} WHERE zipcode = '{zipcode}';";
  tablename = "zipcodes";
  zipcode = "99163";

mysql.query = mysql.query.template;
  mysql.query =  gsub("{tablename}", tablename, mysql.query, fixed=TRUE);
  mysql.query =  gsub("{zipcode}"  , zipcode, mysql.query, fixed=TRUE);

mysql.query;
```

## Connect to DATABASE and RUN QUERY

R drops the database connection quite frequently.  So let's connect and query at the same time.


```{r}
mysql.connection = dbConnect(RMySQL::MySQL(),
                            user       = .db.user,
                            password   = .db.passwd,
                            dbname     = .db.name ,
                            host       = .db.host);

result = dbGetQuery(mysql.connection, mysql.query);
result;
```

## Build a DISTANCE "BOX"

We have the latitude and longitude for PULLMAN, so we can use it to find nearest neighbors.  Many times this is a store within a radius, but we will just look for other zipcodes.  In SQL, we will query in a BOX, and from the results, we will compute distances.

```{r}
radius.miles = 10;

    # let's build a box
    my.latitude = result$latitude[1];
    my.longitude = result$longitude[1];
    
    delta.latitude = radius.miles / 68.703 ;
    delta.longitude = radius.miles / (69.172 * cos(deg2rad(my.longitude))); 
    
    # 4 sides of the square ... CREATE A BOUNDING BOX
    latitude.lower = my.latitude - delta.latitude;
    latitude.upper = my.latitude + delta.latitude;
    
    longitude.lower = my.longitude - delta.longitude;
    longitude.upper = my.longitude + delta.longitude;
```

## Build SQL template
```{r}
mysql.query.template = "SELECT * FROM {tablename} WHERE latitude > {latitude.lower} AND latitude < {latitude.upper} AND longitude < {longitude.lower} AND longitude > {longitude.upper} ORDER BY zipcode ASC;";

mysql.query = mysql.query.template;


mysql.query = gsub("{tablename}",tablename, mysql.query, fixed=TRUE);
mysql.query = gsub("{zipcode}",zipcode, mysql.query, fixed=TRUE);
mysql.query = gsub("{latitude.lower}",latitude.lower, mysql.query, fixed=TRUE);
mysql.query = gsub("{latitude.upper}",latitude.upper, mysql.query, fixed=TRUE);
mysql.query = gsub("{longitude.lower}",longitude.lower, mysql.query, fixed=TRUE);
mysql.query = gsub("{longitude.upper}",longitude.upper, mysql.query, fixed=TRUE);

mysql.query;

```

## Connect to DATABASE and RUN QUERY

R drops the database connection quite frequently.  So let's connect and query at the same time.


```{r}
mysql.connection = dbConnect(RMySQL::MySQL(),
                            user       = .db.user,
                            password   = .db.passwd,
                            dbname     = .db.name ,
                            host       = .db.host);

result = dbGetQuery(mysql.connection, mysql.query);
result;
```

[TODO: how would you take this information and finalize a result?]

# Mapping the results
I have written a function that will map the results of this process.


## My Hometown "Columbia Falls, Montana" `cfalls`

- Find all ZIP codes within 22 miles of Columbia Falls, MT `cfalls` (use lat/long provide from the Wikipedia lookup)... build the bounding "box" and perform the post-hoc "radial distance" computations (as we did in the homework).

```{r}

# copy/paste __student_access__/_SECRET_/_SECRET_database_.txt into console...  or this won't work

# SEVERAL libraries are running under the hood, if you get an error, install them.
# library(maps);
# library(geosphere);     # distm
# library(measurements);  # conv_unit
# library(RMariaDB);
# 
# 
# library(RCurl);
# library(png);

cfalls.latitude = 48.37028; 
cfalls.longitude = -114.18889;
my.radius = 22; my.units = "mi"; #miles

cfalls.info = getNeighborsFromLatLong(22, 48.37028, -114.18889, "mi");

cfalls.info$neighbors;

############## plotting ##############
brown = "#ffe4c4";
green = "#014421";

my.state = "montana";
my.state.color = "#ffe4c4";

my.county = "flathead";
my.county.color = "#014421"; 

my.nearby.states = c("idaho", "washington", "oregon");


plotNeighbors(cfalls.info, 
                    state          = my.state, 
                    state.color    = my.state.color,
                    state.border   = 0.05,
                    county         = my.county, 
                    county.border   = 0.05,  # if you don't see the box, increase this to like 0.75
                    county.color   = my.county.color, 
                    nearby.states  = my.nearby.states); 

```

Why is the "box" not a square?